A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-11-11 thesis/thesis.tex
fix revisions, apply style guidelines

\documentclass[10pt]{article}
\renewcommand{\baselinestretch}{1.8}

@@

\begin{document}
\title{A Wait-free Queue with Poly-logarithmic Step Complexity}
\author{Hossein Naderibeni \\supervised by Eric Ruppert}
\maketitle

\begin{abstract}
In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, the CAS Retry Problem. We show that our algorithm avoids this problem and has a worst-case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
\end{abstract}

\tableofcontents
\pagebreak
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure to work with. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node, and then $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty, it returns \texttt{null}. In a concurrent version of the queue data structure, operations do not happen one at a time. The question that may arise is ``What properties matter for the implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes. Objects like LL/SC, CAS can be used to reach a consensus among any number of processors. A CAS object provides an atomic Compare \& Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}, otherwise it returns \nf{false}.  In this work, we use Compare \& Swap (\nf{CAS}) objects to synchronize among processes.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/good}
  \caption{\label{fig::goodexample}An example of a linearizable execution. Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
\end{figure}

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable. \texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before termination of \texttt{Enqueue(A)} and return \nf{null} or can occur after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
\end{figure}

@@

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps. We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized. Processes can do queries to get information about the linearization stored in the root. We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for the processor. Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case. We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations. Furthermore, we do not address garbage collection but we describe a way we think it can be handled.

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of related work done in the area and the motivation for our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2, we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. We also discuss the motivation and requirements in our design to achieve poly-log time. Section 3.2 contains the algorithm itself.

@@

Chapter 5 is devoted to performance analysis for our implementation. We analyze the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue. In the end, we prove our queue is wait-free.

@@

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues. In the worst-case, step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.4]{pics/msqueue}
  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. In the first diagram, the first element has been dequeued. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

@@

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue. Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket. Each basket has a time interval in which all its nodes' enqueue operations overlap. Since the operations in a basket are concurrent, we can order them in any way. Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations. However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea. At a time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}. To order the operations in a basket, the mechanism in the process algorithm is to \texttt{CAS} again. The successful process will be the next one in the basket, and so on.}
\end{figure}
% \bf{TODO: make small circles the same size}

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue. MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous added node to the new node. They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue. As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one. The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/sundell}
    \caption{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
\end{figure}

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique to achieve wait-freedom. Their step complexity is $\Omega(p)$ because of the helping mechanism.

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures. In their queue, operations  return future objects instead of responses. Later when the response is needed, it can be evaluated from the future object. They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue. They use some properties of the queue size before and after a batch, similar to a part of our work. Their queue is not wait-free, in fact if the batch sizes are 1, then the queue is like MS-queue.

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. One is faster but its running time is not bounded and the other one is slower but has a time-bound. They show that these two paths do not affect each other and the queue remains consistent. If a process makes no progress, other processes help its slow path to finish. The helping phase suffers from the CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the worst-case complexity cannot be better than $\Omega(p)$.

CAS Retry Problem is not limited to list-based queues; array-based queues also share the CAS retry problem as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}. In all implementations where there is a CAS object with contention $p$ ($p$ processes invoking CAS instructions concurrently), CAS Retry Problem exists. Our motivation is to overcome this problem and present a wait-free sublinear queue.

@@

\subsection{Universal Constructions and other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object, introduced by Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$ bit words. Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses using reasonably sized words~\cite{DBLP:conf/podc/ChandraJT98}. Their universal construction cannot be used to create any data structure, but a set of objects called closed objects (a queue is not a closed object). We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions on page 13.

@@

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes. When a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and returns the number of elements in the tree. There are some similarities between designing a queue and a Fetch\&Inc object. A Fetch\&Inc object can be constructed from a queue. The algorithm by Ellen and Woelfel is interesting because it is the first wait-free data structure achieving poly-log implementation after~\cite{DBLP:conf/podc/ChandraJT98}.

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues. Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes. Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is that their result does not contradict ours and we manage to reach poly-log time complexity.

@@

There is a shared binary tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering of the operations invoked by processes. Each process has a leaf in which the operations invoked by the process are stored in order. When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root. Each node of the tree keeps an ordering of operations propagated up to it. All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering. 
\begin{figure}[hbtp]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored. Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}  
\end{figure}

To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to the $n$. We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
We sketch the argument here.

\begin{figure}[hbtp]
\begin{center}
\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{Before the \nf{Refresh}.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{New operations are appended.}
\end{subfigure}
\hspace{6em}
\caption{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append. Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
\end{center}
\end{figure}

\begin{figure}[hbpt]
  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
  \caption{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive \nf{Refresh}es.}
\end{figure}

We use \nf{CAS} (Compare \& Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}). This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the new operations in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended. After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node. So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization. This cooperative solution allows us to overcome the CAS Retry Problem.

>>>>

\documentclass[11pt]{article}
\renewcommand{\baselinestretch}{1.8}

@@

\begin{document}

\begin{center}

\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}

\vspace{2em}
\Large{\bf{Hossein Naderibeni}}

\vspace{3em}
\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science

\vspace{2em}
Graduate Program in Computer Science\\
York University,\\
Toronto, Ontario

\vspace{2em}
November, 2022

\vspace{3em}
\textcopyright Hossein Naderibeni, 2022

\end{center}
\thispagestyle{empty}
\pagebreak

\normalsize
\section*{Abstract}

In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem. We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
\pagenumbering{roman}
\pagebreak

\section*{Acknowledgements}

First, I want to thank my supervisor Eric Ruppert. ...

I would also like to thank professor Franck van Breugel. ...

I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.

I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.

\pagebreak
\tableofcontents

\pagebreak
\listoffigures

\pagebreak
\listoftables
\newpage

\pagenumbering{arabic}
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node, and then $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty, it returns \texttt{null}. In a concurrent version of the queue data structure, operations do not happen one at a time. The question that may arise is ``What properties matter for the implementation of a shared data structure?''. Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy based on how powerful objects are to solve the consensus problem among processes. Objects like LL/SC or CAS can be used to reach a consensus among any number of processors. A CAS object provides an atomic Compare\&Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}; otherwise it returns \nf{false}.  In this work, we use Compare\&Swap (\nf{CAS}) objects to synchronize among processes.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/good}
  \caption[An example of a linearizable execution.]{\label{fig::goodexample}An example of a linearizable execution. Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
\end{figure}

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption[An example of a non-linearizable execution.]{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable. \texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before \texttt{Enqueue(A)} and return \nf{null} or can occur after \texttt{Enqueue(A)} and return \texttt{A}.}
\end{figure}

@@

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps. We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized. Processes can do queries to get information about the linearization stored in the root. We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for the processor. Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case. We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations. Furthermore, we do not address garbage collection but we describe a way we think it can be handled.

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2, we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. We also discuss the motivation and requirements in our design to achieve poly-log time. Section 3.2 contains the algorithm itself and there is an example of a concurrent exuection on the algorithm described in Section 3.3.

@@

Chapter 5 is devoted to complexity analysis for our implementation. We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue. In the end, we prove our queue is wait-free.

@@

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues. In the worst-case, step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or a dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.4]{pics/msqueue}
  \caption[MS-queue structure]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. In the first diagram, the first element has been dequeued. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

@@

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue. Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket. Each basket has a time interval in which all its nodes' enqueue operations overlap. Since the operations in a basket are concurrent, we can order them in any way. Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations. However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the worst-case step complexity remains $\Omega(p)$ per operation.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption[Baskets queue]{At a time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}. To order the operations in a basket, processes perform another \nf{CAS}. The successful process will be the next one in the basket, and so on.}
\end{figure}

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue. MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node. They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue. As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one. The worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/sundell}
    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}]{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
\end{figure}

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom. Their step complexity is $\Omega(p)$ because of the helping mechanism.

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures. In their queue, operations  return future objects instead of responses. Later when the response is needed, it can be evaluated from the future object. They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue. They use some properties of the queue size before and after a batch, similar to a part of our work. Their queue is not wait-free: in fact, if the batch sizes are 1, then the queue is like MS-queue.

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. The fast path  ensures good performance while the slow path ensures termination. They show that these two paths do not affect each other and the queue remains consistent. If a process makes no progress, other processes help its slow path to finish. The helping phase suffers from the CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the worst-case complexity cannot be better than $\Omega(p)$.

The CAS Retry Problem is not limited to list-based queues; array-based queues also share it as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  Our motivation is to overcome this problem and present a wait-free sublinear queue.

@@

\subsection{Universal Constructions and Other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words. Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. However their algorithm cannot be used to create a queue. We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of the page 13.

@@

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes. When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree. There are some similarities between designing a queue and a Fetch\&Inc object. A Fetch\&Inc object can be constructed from a queue. The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues. Also, it is one of the few wait-free data structures achieving poly-log complexity.

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues. Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes. Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is that since their result is about when contention is low, their lower bound does not contradict with our algorithm's complexity and we manage to reach poly-log time complexity.

@@

There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes. The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree})Each process has a leaf in which the operations invoked by the process are stored in order. When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root. Each node of the tree keeps an ordering of operations propagated up to it. All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering. 
\begin{figure}[hbtp]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption[Tournament tree]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored. Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}  
\end{figure}

To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$. We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
We sketch the argument here.

\begin{figure}[hbtp]
\begin{center}
\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{Before the \nf{Refresh}}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{After the \nf{Refresh}}
\end{subfigure}
\hspace{6em}
\caption[A successful \nf{Refresh}]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append. Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
\end{center}
\end{figure}

\begin{figure}[hbpt]
  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
  \caption[Two consecutive failed \nf{Refresh}es by a process]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
\end{figure}

We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}). This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended. After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node. So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization. This cooperative solution allows us to overcome the CAS Retry Problem.

It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
Instead we use an implicit representation of the operations propagated together. Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}). Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block stores information about the operations propagated by a \nf{Refresh}. It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure. See Figure~\ref{fig:block} for an example. A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent. We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently. Thus, there are at most $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.
